### 2021-05-11 proposal/proposal.Rmd
updated proposal

Graph and network data have become increasingly more widespread in various fields including sociology, neuroscience, biostatistics, and computer science. This has resulted in certain challenges for researchers, as most traditional statistical and machine learning methods are incompatible with graphs and instead require the data to exist in a subset of Euclidean space. This includes clustering and community detection, which are the typical problems of interest when working with graphs. Common clustering methods often involve calculating some central or representative point around which the data belonging to that cluster lie (e.g., Lloyd's algorithm for $K$-means clustering \cite{1056489}, Gaussian Mixture Models \cite{doi:10.1198/016214502760047131}). Because these methods involve computing summary statistics within each cluster, they cannot be applied directly to graphs as community detection algorithms.

Some proposed methods for unifying graph community detection with traditional clustering techniques involve embedding the graph into Euclidean space, of which the most popular is Spectral Clustering \cite{DBLP:journals/corr/abs-0711-0189}. The Random Dot Product Graph (RDPG) \cite{10.1007/978-3-540-77004-6_11} and Generalized Random Dot Product Graph (GRDPG) \cite{rubindelanchy2017statistical} models take this further by explicitly constructing generative models such that latent positions in Euclidean space are used to induce graphs. A community detection method motivated by this would involve learning the latent positions given an observed graph and then learning the community labels given the latent positions.

@@

Let $G = (V, E)$ be an undirected, unweighted graph with no self-loops with $n$ vertices. Denote $A \in \{0, 1\}^{n \times n}$ as the adjacency matrix for $G$ such that $A_{ij} = 1$ if there exists an edge between vertices $i$ and $j$ and $A_{ij} = 0$ otherwise. Because $G$ is symmetric and contains no self-loops, $A_{ij} = A_{ji}$ and $A_{ii} = 0$ for $i, j \in [n]$. We further restrict our analyses to independent Bernoulli graphs. Let $P \in [0, 1]^{n \times n}$ be a symmetric matrix of edge probabilities. Graph $G$ is sampled by $A_{ij} \indep P_{ij}$ for each $0 \leq i < j \leq n$ ($A_{ji} = A_{ij}$ and $A_{ii} = 0$). Denote $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top \in \mathbb{R}^{n \times d}$ as the sample $x_1, ..., x_n \in \mathbb{R}^d$, and denote $z_1, ..., z_n \in [K]$ as their corresponding (hidden) labels.

@@

Generative models for symmetric Bernoulli graphs involve defining the edge probability matrix $P$ whose $ij$^th^ entry is the probability of an edge between vertices $i$ and $j$ for each $i, j \in [n]$. Community detection and recovery starts by restricting the generative model such that for each pair of vertices, the probability of an edge between the vertices is conditioned on the labels of the vertices. One such model is the Stochastic Block Model (SBM) \cite{doi:10.1080/0022250X.1971.9989788}: Given $K$ communities and each vertex belonging to one community, the SBM restricts $P$ to $K (K + 1) / 2$ unique entries such that $P_{ij} = B_{z_i z_j}$ where $B \in [0, 1]^{K \times K}$ with entries $B_{kl}$ denoting the edge probability of each vertex in community $k$ having an edge with each vertex in community $l$. The homogeneous SBM further restricts $P$ to two unique entries such that $P_{ij} = p$ if $z_i = z_j$ and $P_{ij} = q$ otherwise. Multiple generalizations of the SBM have been introduced since, including the Degree Corrected Block Model (DCBM) and the Popularity Adjusted Block Model (PABM). Like the SBM, these models involve edge probability matrix $P$ that is restricted to rank $d < n$ in part based on community labels.

@@

The RDPG starts with points in latent space $X \in \mathcal{X} \subset \mathbb{R}^d$ such that $\forall x, y \in \mathcal{X}$, $x^\top y \in [0, 1]$. $P$ is then constructed as $P = X X^\top$, and graph $G$ with adjacency matrix $A$ is drawn from $P$. We provide a more formal definition of the RDPG and GRDPG below.

@@

The structure of the RDPG and GRDPG provides an intuitive method for recovery of the latent positions via spectral embedding.

@@

\citeauthor{athreya2017statistical} showed that under mild conditions, if $(A_n, X_n) \sim \text{RDPG}(F, n)$ and $\hat{X}_n$ is the scaled ASE of $A_n$, for some sequence of orthogonal matrices $W_n$,

@@

It is straightforward to show that all Bernoulli graphs with positive semidefinite $P$ are special cases of the RDPG, which is a special case of the GRDPG, and all graphs generated by $P$ are special cases of the GRDPG. This includes the SBM, DCBM, and PABM.

@@

## Subspace Clustering

Subspace clustering, which assumes that points in $\mathbb{R}^d$ each lie on one of $K$ subspaces of $\mathbb{R}^d$, is an approach that has found a wide range of uses by the Statistics and Machine Learning communities, particularly within the field of Computer Vision \cite{5206547}.

Of particular interest is Sparse Subspace Clustering (SSC), which is performed by solving an optimization problem for each observed point in a sample. Given $X \in \mathbb{R}^{n \times d}$ with vectors $x_i^\top \in \mathbb{R}^d$ as rows of $X$, the optimization problem $c_i = \min\limits_{c} \|c\|_1$ subject to $x_i = X_{-i} c$ and $c^{(i)} = 0$ is solved for each $i \in [n]$. The solutions are collected into matrix $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}^\top$ to construct affinity matrix $B = |C| + |C^\top|$. If each $x_i$ lie perfectly on one of $K$ subspaces, $B$ is sparse such that $B_{ij} = 0$ $\forall x_i, x_j$ belonging to different subspaces. Then $B$ can describe a graph with at least $K$ disjoint subgraphs, and if the number of subgraphs is exactly $K$, each subgraph maps onto a subspace.

@@

As discussed in \S 2.2, if the latent positions of a RDPG or GRDPG model are such that they lie on a small number of subspaces, ASE followed by SSC may be able to identify whether vertices of the graph $v_i, v_j$ belong to the same subspace or to different subspaces, and one such model that is consistent with this construction is the PABM.

@@

We will first define the PABM.

@@

This leads to the following theorem.

@@

If $\hat{V}$ is the *unscaled* ASE of $A$, Theorem \ref{osc} and results from  \citeauthor{rubindelanchy2017statistical} together imply $n \hat{V} \hat{V}^\top \stackrel{a.s.}{\to} 0$, leading to the following result:

@@

@@

## Manifold Clustering

We would like to extend subspace clustering to manifold clustering. The problem setup is as follows: Suppose that in the latent space $\mathcal{X} \subset \mathbb{R}^d$, sample $X$ of $n$ points lie on a union of $K$ disjoint manifolds with each manifold corresponding to a community. If $A \sim \text{RDPG}(X)$, we wish to recover the community labels (up to permutation) from $A$.

Similarly, suppose that probability distribution $F$ is described as follows:

1. Define functions $f_1, ..., f_K$ such that 
$f_k : [0, 1] \mapsto \mathcal{X}$ and $f_k(t) \neq f_l(t)$ $\forall k, l \in [K]$ 
and $t \in [0, 1]$.
2. Sample labels
$z_1, ..., z_n \stackrel{iid}{\sim} Categorical(\pi_1, ..., \pi_K)$.
3. Sample $t_1, ..., t_n \stackrel{iid}{\sim} D$ where $D$ has support 
$[0, 1]$. 
4. Set latent positions $x_i = f_{z_i}(t_i)$ and 
$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.

@@

### Affine Subspaces

@@

If instead of sampling uniformly from line segments of unit length, we sample uniformly from a 1 dimensional manifolds of unit length, the above property still holds. Let $U_1, ..., U_n \stackrel{iid}{\sim} Uniform(0, 1)$ and $f : [0, 1] \mapsto \mathbb{R}^d$ be a smooth function such that $\int_u^v \sqrt{1 + (f'(t))^2} dt = \|u - v\|$. Then $U_{(i+1)} - U_{(i)} \geq \|f(U_{(i+1)}) - f(U_{(i)})\|$, so $P(U_{(i+1)} - U_{(i)} \leq \alpha) \leq P(\|f(U_{(i+1)}) - f(U_{(i)})\| \leq a)$. If the shortest distance between the two manifolds defined by $f_1$ and $f_2$ with the same restriction is $a$, then the same $N$ as before is sufficient, although perhaps a more lenient lower bound can be derived based on the shape of $f_i(\cdot)$.

>>>>

Graph and network data have become increasingly widespread in various fields including sociology, neuroscience, biostatistics, and computer science. This has resulted in various challenges for researchers who rely on traditional statistical and machine learning methods, many of which are incompatible with graph data and instead require the data to exist as feature vectors in Euclidean space. Such challenges often involve clustering and community detection. Common clustering methods typically involve calculating some central or representative point for each cluster around which the data belonging to that cluster lie (e.g., Lloyd's algorithm for $K$-means clustering \cite{1056489}, Gaussian Mixture Models \cite{doi:10.1198/016214502760047131}). Because these methods involve computing summary statistics within each cluster, such as the sample average, they cannot be applied directly to graphs, necessitating methods for transforming the graph data into feature vectors.

One family of methods for unifying graph community detection with traditional clustering techniques is Spectral Clustering \cite{DBLP:journals/corr/abs-0711-0189}, which involves embedding the graph into Euclidean space, followed by applying a popular clustering algorithm such as $K$-means clustering. The Random Dot Product Graph (RDPG) \cite{10.1007/978-3-540-77004-6_11} and Generalized Random Dot Product Graph (GRDPG) \cite{rubindelanchy2017statistical} models take this further by explicitly constructing generative models such that latent positions in Euclidean space are used to induce graphs. A community detection algorithm motivated by this may involve learning the latent positions given an observed graph and then learning the community labels given the latent positions.

@@

Let $G = (V, E)$ be an undirected, unweighted graph with no self-loops with $n$ vertices. Denote $A \in \{0, 1\}^{n \times n}$ as the adjacency matrix for $G$ such that $A_{ij} = 1$ if there exists an edge between vertices $i$ and $j$ and $A_{ij} = 0$ otherwise. Because $G$ is symmetric and contains no self-loops, $A_{ij} = A_{ji}$ and $A_{ii} = 0$ for $i, j \in [n]$. We further restrict our analyses to independent Bernoulli graphs. Let $P \in [0, 1]^{n \times n}$ be a symmetric matrix of edge probabilities. Graph $G$ is sampled from $P$ by drawing $A_{ij} \indep \Bernoulli(P_{ij})$ for each $0 \leq i < j \leq n$ ($A_{ji} = A_{ij}$ and $A_{ii} = 0$). Finally, we denote $X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top \in \mathbb{R}^{n \times d}$ as the sample $x_1, ..., x_n \in \mathbb{R}^d$, and denote $z_1, ..., z_n \in [K]$ as their corresponding (hidden) labels.

@@

Generative models for symmetric Bernoulli graphs involve defining the edge probability matrix $P$ whose $ij$^th^ entry is the probability of an edge between vertices $i$ and $j$ for each $i, j \in [n]$. In order to motivate community detection methods, we restrict the generative model such that for each pair of vertices, the probability of an edge between the vertices is conditioned on the labels of the vertices. One such model is the Stochastic Block Model (SBM) \cite{doi:10.1080/0022250X.1971.9989788}: Given $K$ communities and each vertex belonging to one community, the SBM restricts $P$ to $K (K + 1) / 2$ unique entries such that $P_{ij} = B_{z_i z_j}$ and $B \in [0, 1]^{K \times K}$ with entries $B_{kl}$ denoting the edge probability of each vertex in community $k$ having an edge with each vertex in community $l$. The homogeneous SBM further restricts $P$ to two unique entries such that $P_{ij} = p$ if $z_i = z_j$ and $P_{ij} = q$ otherwise. Multiple generalizations of the SBM have been introduced since, including the Degree Corrected Block Model (DCBM) and the Popularity Adjusted Block Model (PABM). Like the SBM, these models involve edge probability matrix $P$ that is restricted to rank $d < n$ in part based on community labels.

@@

Like the SBM, DCBM, and PABM, the RDPG and GRDPG are generative models for graphs involving an edge probability matrix $P$. The RDPG starts with points in latent space $X \in \mathcal{X} \subset \mathbb{R}^d$ such that $\forall x, y \in \mathcal{X}$, $x^\top y \in [0, 1]$. $P$ is then constructed as $P = X X^\top$, and graph $G$ with adjacency matrix $A$ is drawn from $P$. We provide a more formal definition of the RDPG and GRDPG below.

@@

The structure of the RDPG and GRDPG provides a straightforward method for recovery of the latent positions via spectral embedding.

@@

\citet{athreya2017statistical} showed that under mild conditions, if $(A_n, X_n) \sim \text{RDPG}(F, n)$ and $\hat{X}_n$ is the scaled ASE of $A_n$, for some sequence of orthogonal matrices $W_n$,

@@

It is straightforward to show that all Bernoulli graphs with positive semidefinite $P$ are special cases of the RDPG, which is a special case of the GRDPG, and all graphs generated by $P$ are special cases of the GRDPG. This includes the SBM, DCBM, and PABM. In the following example, we show that the ASE of the SBM has a very particular form.

@@

@@

As discussed in \S 2.1, all Bernoulli graph models can be expressed as a RDPG or GRDPG model, and in particular, the (associative) SBM is a RDPG with a very specific structure in the latent space. A similar result has been shown for the DCBM. We will now extend this to the PABM.

@@

We first define the PABM.

@@

In Theorem \ref{theorem2}, we showed that a possible latent configuration for a $K$-community PABM under the GRDPG framework consists of latent positions in $\mathbb{R}^{K^2}$ such that each community corresponds to a $K$-dimensional subspace such that each subspace is orthogonal to the others.

@@

If $\hat{V}$ is the *unscaled* ASE of $A$, then results from  \citeauthor{rubindelanchy2017statistical} imply $n \hat{V} \hat{V}^\top \stackrel{a.s.}{\to} n V V^\top$. Then for vertices $i$ and $j$ belonging to separate communities, the $ij$^th^ entry of $n \hat{V} \hat{V}^\top$ approaches 0 with probability 1:

@@

In addition to OSC, since the communities are subspaces in the latent space, we can leverage existing methods for subspace clustering. Of particular interest is Sparse Subspace Clustering (SSC), which is performed by solving an optimization problem for each observed point in a sample. Given $X \in \mathbb{R}^{n \times d}$ with vectors $x_i^\top \in \mathbb{R}^d$ as rows of $X$, the optimization problem $c_i = \min\limits_{c} \|c\|_1$ subject to $x_i = X_{-i} c$ and $c^{(i)} = 0$ is solved for each $i \in [n]$. The solutions are collected into matrix $C = \begin{bmatrix} c_1 & \cdots & c_n \end{bmatrix}^\top$ to construct affinity matrix $B = |C| + |C^\top|$. If each $x_i$ lie perfectly on one of $K$ subspaces, $B$ is sparse such that $B_{ij} = 0$ $\forall x_i, x_j$ belonging to different subspaces. Then $B$ can describe a graph with at least $K$ disjoint subgraphs, and if the number of subgraphs is exactly $K$, each subgraph maps onto a subspace.

@@

## Proposed Generative Graph Models

In the previous sections, we connected well-known and highly structured generative graph models to the RDPG and GRDPG to show that highly structured latent configurations generate graphs consistent with these models: The latent space for the SBM consists of $K$ point masses, the latent space for the DCBM consists of $K$ rays emanating from the origin, and the latent space for the PABM consists of $K$ $K$-dimensional subspaces. In the following sections, we explore additional structured latent configurations corresponding to community structure and develop methods for community detection based on the consistency of the ASE and the structural forms of the latent configurations.

The general structure of interest can be described as follows: Suppose that in the latent space $\mathcal{X} \subset \mathbb{R}^d$, sample $X$ of $n$ points lie on a union of $K$ disjoint manifolds with each manifold corresponding to a community. If $A \sim \text{RDPG}(X)$, we wish to recover the community labels (up to permutation) from $A$.

Equivalently, suppose that probability distribution $F$ is described as follows:

1. Define functions $f_1, ..., f_K$ such that 
$f_k : [0, 1] \mapsto \mathcal{X}$ and $f_k(t) \neq f_l(t)$ $\forall k, l \in [K]$ 
and $t \in [0, 1]$.
2. Sample labels
$z_1, ..., z_n \iid Categorical(\pi_1, ..., \pi_K)$.
3. Sample $t_1, ..., t_n \iid D$ where $D$ has support 
$[0, 1]$. 
4. Set latent positions $x_i = f_{z_i}(t_i)$ and 
$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.

@@

### Affine Segments

@@

If instead of sampling uniformly from line segments of unit length, we sample uniformly from a 1 dimensional manifolds of unit length, the above property still holds. Let $U_1, ..., U_n \stackrel{iid}{\sim} Uniform(0, 1)$ and $f : [0, 1] \mapsto \mathbb{R}^d$ be a smooth function such that $\int_u^v \sqrt{\sum_k (df_k / dt)^2} dt = \|u - v\|$. Then $U_{(i+1)} - U_{(i)} \geq \|f(U_{(i+1)}) - f(U_{(i)})\|$, so $P(U_{(i+1)} - U_{(i)} \leq \alpha) \leq P(\|f(U_{(i+1)}) - f(U_{(i)})\| \leq a)$. If the shortest distance between the two manifolds defined by $f_1$ and $f_2$ with the same restriction is $a$, then the same $N$ as before is sufficient, although perhaps a more lenient lower bound can be derived based on the shape of $f_k(\cdot)$.
